PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-19 peace/papers/NAR_web_server_2009/paper.sup.tex
Modifications to paper; addition of natbibOrig.sty file

@@

% Using the following to mark changes for review
\usepackage{color}
\definecolor{blue}{rgb}{0,0,1}
\newcommand{\mc}[1]{\textcolor{blue}{#1}}
%\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring

@@

In the above discussion of $d^2$ we assume sequences of at least 100 bp, and
in pratice use a window of size 100.  While the technique should be,
in principle, adaptable to short-read sequences (decreasing window
size as necessary), there are some problems in the details.
Diminishing window sizes implies both diminishing accuracy (requiring
the thrshold be adjusted) and increasing runtime (as there are
potenitally more window pairs).  More importantly, there is the
problem of non-homogonous sequence lengths: short-read technologies
tend to produce short reads, but can still include quite long reads in
their output.  Hence it is not immediately clear how to compare a very
short read against a very long one.

We address these issues with our {\it adaptive $d^2$} strategy that
works as follows: we put sequences, based on length, into three {\it
  overlapping} groups (i.e. sequences less than length XXX, sequences
between length YYY and ZZZ, and seqeunces greater than AAA).  For each
group we pick a constant windowsize and threshold tailored to the
length range of the group (as well as filtering heurstic constants),
and use those values when computing the $d^2$ distance for two
sequences in the same group.  We do not compute distances between
sequences in different groups. Once computed, we assign to the MST
edge the ratio of the $d^2$ score to the group threshold, thus
normalizing the weight for comparisons ``between groups'' and
effectivally giving us an edge-removal threhold of 1.

@@

In the \textsc{peace} implementation, we subdevide into three groups:
NEED INFO.

@@

\mc{ Our primary measurement tools if mesurement are sensitivity, the
Jaccard Index, Type 1 error and Type 2 error
(Figure~\ref{sensitivityFig}).  Sensitivity is the ratio for true
positives to actual positives ($tp / (tp + fn)$) -- the fraction of
sequences from the same transcript that are identified as such.  The
{\it Jaccard} index combines both sensitivity and specificity,
calculated as the ratio of true positives to the sum of true
positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
[\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes
before generating the simulated EST sets, we found all three tools to
have 100\% specificity: no false positives. }

\mc{Type 1 and Type 2 errors, as defined in {\it Wang et al.}
[\cite{Wang04}], measure quality at the level of the genes from which
the ESTs were derived.  For Type 1 error, we look at the fraction of genes
that were broken into two or more partitions, while in
Type 2 error we look at the fraction of clusters that contain two or
more genes.   }

@@

\subsection{Raising the Threshold}
\mc{ 
  The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
  two sequences are in the same cluster.  As this is equivilent to our
  $T$ threshold for dismissing MST edges, we have set it to the same
  default value when used for Sanger sequences.  However, we did
  experiment with higher values, to interesting results.  A basic
  analysis indicated that a $T$ value as high as 130 should not have a
  significant impact on its effectivness as a descriminator.  And in
  fact this was born out in simulation, as shown in
  Figure~\label{SiJiT1T2130}.  In contrast with the equivilent results
  for a threshold of $T=40$, as shown in the manuscript, we see
  considerable improvement in \peace\/ over \wcd\/ in all metrics
  except Type 2 error, in which \wcd\/ slightly
  better. Unsurprisingly, this also makes a significant difference in
  the tool's ability to distinguish duplications: we see fairly poor
  performance in Figure~\ref{dups130}.  
}

\mc{
  Most importantly, while the higher threshold appeared to be better
  in theory and in analysis, it did not hold in practice.  In
  Table~\ref{SangerTable130} we show the match to the manuscript
  Table~\ref{SangerTable}, and observe the poor performance of
  \peace\/ when looking at the Jaccard index.  It is clear that in the
  real data, the level of similarity is consierably higher then in the
  simulations, thus making the reduced specificity of the higher
  threshold considerably more of a problem.
XXX}

@@

\section{Raising the Threshold Real Data Sets}

\begin{table}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular} \\
\hline EasyCluster Human & \peace & 0.999 & 0.106 & 0.027 & 0.030 & 67 & 2 & 295 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.937 & 0.345 & 0.337 & 0.0342 & 18256 & 8135 & 1293 \\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\label{SangerReal130}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
  standard quality measurments with $T=130$.}
\end{table}

>>>>

%\usepackage[pdftex,
%colorlinks=true, %makes the links show up in color, rather than in a box
%citecolor=black, %color of the in-text citation numbers
%urlcolor=blue %color of the url links
%]{hyperref}

@@

% Using the following to mark changes for review
\usepackage{color}
\definecolor{red}{rgb}{0,0,1}
\newcommand{\mc}[1]{\textcolor{red}{#1}}
%\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring

@@

In the above discussion of $d^2$ we assume sequences of at least 100
bp, and in pratice use a window of size 100.  While the technique
should be, in principle, adaptable to short-read sequences (decreasing
window size as necessary), there are some problems in the details.
Diminishing window sizes implies both diminishing accuracy (requiring
the thrshold be adjusted) and increasing runtime (as there are
potenitally more window pairs).  More importantly, there is the
problem of non-homogonous sequence lengths: short-read technologies
tend to produce short reads, but some technologiges can still include
long reads in their output.  Hence it is not immediately clear how to
compare a very short read against a very long one.

We address these issues with our {\it adaptive $d^2$} strategy that
works as follows: we partition sequences by length into disjoint groups.  For each
group we pick a constant windowsize and threshold tailored to the
length range of the group (as well as filtering heurstic constants),
and use those values when computing the $d^2$ distance for two
sequences in the same group.  We do not compute distances between
sequences in non-adjaecent groups (i.e. a member of the ``shortest
groups'' against a member of the ``longest groups'' if there is an
interviening sized group).  Once computed, we assign to the MST
edge the ratio of the $d^2$ score to the group threshold, thus
normalizing the weight for comparisons ``between groups'' and
effectivally giving us an edge-removal threhold of 1.

@@

In the \textsc{peace} implementation, we subdevide into three groups:
\begin{itemize}
\item Less than 150 ($w=50$, $T=40$, $t=20$, $u=4$).
\item Between 150 and 400 ($w=75$, $T=60$, $t=35$, $u=6$)
\item Greater than 400 ($w=100$, $T=80$, $t=50$, $t=50$).
\end{itemize}

@@

\mc{Our primary measurement tools if mesurement are sensitivity, the
Jaccard Index, Type 1 error and Type 2 error
(Figure~\ref{sensitivityFig}).  Sensitivity is the ratio for true
positives to actual positives ($tp / (tp + fn)$) -- the fraction of
sequences from the same transcript that are identified as such.  The
{\it Jaccard} index combines both sensitivity and specificity,
calculated as the ratio of true positives to the sum of true
positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
[\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes
before generating the simulated EST sets, we found all three tools to
have 100\% specificity: no false positives.}

\mc{Type 1 and Type 2 errors, as defined in {\it Wang et al.}
[\cite{Wang04}], measure quality at the level of the genes from which
the ESTs were derived.  For Type 1 error, we look at the fraction of genes
that were broken into two or more partitions, while in
Type 2 error we look at the fraction of clusters that contain two or
more genes.}

@@

\subsection{Raising the Threshold}
\label{sim_thresh}
\mc{ 
  The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
  two sequences are in the same cluster.  As this is equivilent to our
  $T$ threshold for dismissing MST edges, we have set it to the same
  default value when used for Sanger sequences.  However, we did
  experiment with higher values, to interesting results.  A basic
  analysis indicated that a $T$ value as high as 130 should not have a
  significant impact on its effectivness as a descriminator.  And in
  fact this was born out in simulation, as shown in
  Figure~\label{SiJiT1T2130}.  In contrast with the equivilent results
  for a threshold of $T=40$, as shown in the manuscript, we see
  considerable improvement in \peace\/ over \wcd\/ in all metrics
  except Type 2 error, in which \wcd\/ slightly
  better. Unsurprisingly, this also makes a significant difference in
  the tool's ability to distinguish duplications: we see fairly poor
  performance in Figure~\ref{dups130}.  
}

\mc{
  Most importantly, while the higher threshold appeared to be better
  in theory and in analysis, it did not hold in practice.  In
  Table~\ref{SangerTable130} we show the match to the manuscript
  Table~\ref{SangerTable}, and observe the poor performance of
  \peace\/ when looking at the Jaccard index.  It is clear that in the
  real data, the level of similarity is consierably higher then in the
  simulations, thus making the reduced specificity of the higher
  threshold considerably more of a problem.
}

@@

\section{Real Data}

\subsection{Runtimes}

In Table~\ref{RunTime} we show the runtime of \peace\/ and \wcd\/ on
five different benckmark sets obtained from the WCD and EasyCluster
studies [\cite{Hazelhurst08a,Picardi09}].

\begin{table}[tbp]
\begin{center}
\begin{tabular} {|c|| c | c || c | c |}
\hline
 Bencmark & \begin{tabular}{c} Number of \\ Sequences \end{tabular}
 & \begin{tabular}{c} Number of \\ Bases \end{tabular} & Peace Runtme (s)&
 WCD Runtime (s) \\
\hline
\hline
 Human Cluster & 17733 & 11.7 Mb & 293 & 804 \\
\hline
 Cotton & 29,995 & 17.1 Mb & 222 & 339 \\
\hline
 Ricinus & 57,690 & 40.9 Mb & 1344 & 2356 \\
\hline
 A076941 & 76,941 & 32.8 Mb & 1166 & 966 \\
\hline
\hline
 A686903 & 686,904 & 294.7 Mb & 9449 & 2481 \\
\hline
\end{tabular}
\end{center}
\caption{Runtime for \peace\/ and \wcd\/ on benchmarks from the
  WCD and EasyCluster studies [\cite{Hazelhurst08a, Picardi09}].
  Runtimes for the first four sets are sequential, while that last was
  run over thirty processors.}\label{RunTime}
\end{table}

\subsection{Higer Threshold}

In Table~\ref{SangerReal130} we see the effect on the application of
\peace\/ to the data sets from Table~\ref{SangerReal} when raising the
threshold $T$ to 130 (see pervious discussion in
Section~\ref{sim_thresh}.  While doing this appeared to be fairly
effective in simulation, it did not hold up in pracice.  We see
essentially no change in the sensitivity, but significant changes in
the Jaccard Index (and a corresponding decrease in the total number of
clusters).  We suspect that this is due to a higher number of more
closely related genes than are present in the simulated data.  We can
conclude that, without considerably more investigation, we should be
settig $T$ to 40.

\begin{table}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime
      (s) \end{tabular} \\
\hline
\hline EasyCluster Human & \peace & 0.999 & 0.106 & 0.027 & 0.030 & 67 & 2 & 295 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.937 & 0.345 & 0.337 & 0.0342 & 18256 & 8135 & 1293 \\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\label{SangerReal130}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
  standard quality measurments with $T=130$.}
\end{table}
